Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP

### 2020-11-16 paper/Paper.tex
updates

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
        T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
    \title{Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP}
    \author{\IEEEauthorblockN{Benjamin Pierce\\\href{mailto:bgp12@case.edu}{bgp12@case.edu}}
        \IEEEauthorblockA{
            \textit{Department of Computer and Data Sciences}\\ }
        \and
        \IEEEauthorblockN{Alberto Safra\\\href{mailto:axs1249@case.edu}{axs1249@case.edu}}
        \IEEEauthorblockA{ 
	    \textit{NCDP, Dean's Deptartment}}
        \and
        \IEEEauthorblockN{Colin Causey\\\href{mailto:cjc158@case.edu}{cjc158@case.edu}}
        \IEEEauthorblockA{ 
            \textit{Department of Computer and Data Sciences}}
        \and
        \IEEEauthorblockN{Jason Richards\\\href{mailto:jdr145@case.edu}{jdr145@case.edu}}
        \IEEEauthorblockA{
            \textit{Department of Computer and Data Sciences}}}
    \maketitle

@@

    As mentioned before, all comparison based sorts have an asymptotic lower bound of $\Omega(nlog(n)$. 
    Indeed,  Mergesort, and Heapsort achieve a $O(nlog(n))$ upper bound as well, and Quicksort behaves as $O(nlog(n))$ in the average case. 
    Insertion sort, as a less advanced algorithm, has a upper bound of $O(n^2)$, as does Quicksort. \cite{cormen_introduction_2009} 
    However, Quicksort's upper bound is rarely the case, and tends to have smaller constant factors then either Mergesort or Heapsort. \cite{hoare_algorithm_1961} % TODO wrong cite
    Additionally, each algorithm is constructed differently, and some are more amendable to parallelization than others.

@@

    \subsection{Heapsort}
    The final sort dicussed here is Heapsort. Heapsort is yet another $O(n log(n))$ comparison sort. 
    It, along with the heap data structure, was invented in 1964. \cite{forsythe_algorithms_1964}
    Heapsort first turns the dataset into a max heap, which is a binary tree where each parent node is greater then its children. 
    This process, called \textit{heapification}, is an $O(n)$ algorithum. 
    Sorting is performed by repeatably popping the root node (the maximum value) and re-heapifying. 
    This procedure takes advantage of the binary tree structure, and is worst case $O(n log (n))$ overall.  \cite{cormen_introduction_2009}
    Unfortunately, Heapsort is a poor candidate for parallelization, as it does not divide or partition the input array into subarrays and depends on the root node being the absolute maximum.

@@

    In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. \cite{blelloch_programming_1996}
    This is because each subarray can be sorted independently, and this leads to speedup. 
    % TODO @Jason add detail on your implementation
    \section{Methodology}
    In this study, the OpenMP \cite{openmp08} API is used for parallization. 
    It is an example of a fork-join methodology	, where each parallel thread forks off from a main thread, then, the results are joined back together. 
    This is implemented into the C programming language via \textit{\#pragma} preprocessor directives. 
    OpenMP is useful for obviously parallel cases, such as Mergesort and Quicksort, as the division of work is quite clear. 
    % TODO add more OMP detail here.

@@

@@

    \subsection{Quicksort}
    There was great success with parallel Quicksort. 
    As a divide-and-conquer algorithm, it is a naturally parallel algorithum. 
    The results of this algorithum can be seen in Figure \ref{qck_per}. 
    \begin{figure}[h]
        \includegraphics[width=9.25cm]{qs_mt.png} 
        \caption{Performance of Quicksort}
        \label{qck_per}
    \end{figure}

@@

    \begin{algorithm}
        \SetAlgoLined
        \KwResult{A sorted list }
        list $a$\;
        list $result$\;
        int $threads$\;
        let $lists$ be $a$ partitioned into $threads$\\
        \While{every list in lists is not empty }{
            \textbf{In Parallel}\;
            \hspace{0.5cm}$heapify$ each list\;
            $pop$ the largest element of all sub-heaps into $result$\;
            $re-heapify$ the heap that has been popped\;
        }
        \Return $result$
        \caption{Parallel-Heaps}
        \label{alg}
    \end{algorithm}
    Algorithm \ref{alg} presents a possibly parallel algorithm that utilizes the idea of Heapsort. 
    Essentially, the algorithm divides the array into many smaller heaps, each managed by a single thread. 
    This way, all heapification can be done in parallel, and the "max" item is selected from the number of heaps through a straightforward traversal of the roots of the sub-heaps. 
    However, this algorithum has several downsides. 
    One is that the end reheapification is still sequential; as by necessity, one heap must be popped, the algorithm must wait on the reheapification of that heap. 
    This heap will be smaller by a constant factor then the traditional, single-threaded Heapsort, however. 
    Therefore, $Parallel-Heaps$ will be at most a constant speedup for Heapsort.

@@

    For our analysis, we attempted to implement an additional algorithm, Shellsort. We were unsuccessful. We include Shellsort here, however, for completeness and for the Case Western Reserve University HPC developer that may find themselves wondering about Shellsort's parallel implementation. Shellsort is an seen to be a generalization of a more elementary sorting algorithm, Insertionsort, and uses the Insertionsort algorithm in its subroutines. Because of Insertionsort's dependence on access to the entire input array per swap movement, all Insertionsort swaps must be done in serial or risk a data race, which would jepordize the sort's integrity. For this reason, Insertionsort is nonparallelizable, yet performant in small array size scenarios. Shellsort, however, improves on Insertionsort's design by incorporating a divide and conquer strategy. In particular, Shellsort begins by in-place Insertionsorting small, distant subsets of the input array before peforming Insertionsort on the whole resulting array. This final Insertionsort is done because Insertionsort has fantastic $O(N)$ performance for almost-sorted arrays.  Since all Shellsort's subroutines contribute to advancing the array's state in regards to reaching an optimal setup for the final Insertionsort call and that Shellsort's final serial Insertionsort call on the whole array will inevitably fix any gaps the setup stage missed, Shellsort is a natural candidate to parallelization. A perfunctory parallelization attempt, like the one attemped in the appendix, then can implemented by partitioning an input array into disjoint blocks, having sorting them each sorted in parallel with an OMP recursive call to Shellsort, and then running Shellsort in serial on the result of all those in-place recursive sorts. Note Shellsort's particular potential success in giving each CPU of the OMP tasks significant work to do per unit parallel time. This is what would contribute to such an Implentations potential average span of $O(logn)$. Unfortunately, such an implementation was not found to be possible for us at this time, due to issues with redistrubuting the number of active cores in a loop using the OpenMP framework.

    \subsection{Comparative Results}
    Results are compared in a few different ways. 
    One is to hold the number of threads constant, and plot the performance of each algorithm, as in Figure \ref{msqshs}, which shows that Heapsort, which is not paralleized, proves to be much worse then either Quicksort or Mergesort.
    \begin{figure}
        \includegraphics[width=9.25cm]{mshsqs.png} 
        \caption{Comparative Performance with 8 threads}
        \label{msqshs}
    \end{figure}
    %TODO Add deets
    Additionally, it is useful to compare results on a single array size with multiple threads. 
    This can be seen for $10^6$ elements in Figure \ref{1e5}
    \begin{figure}
        \includegraphics[width=9cm]{1e5.png} 
        \caption{Comparative Performance with for $10^4$ elements}
        \label{1e5}
    \end{figure}
    Interestingly, the performance of Quicksort appears to worsen with the number of threads more then Mergesort; this suggests significant overhead introduced by parallelization. 
    However, this effect goes away with an increase in the number of elements, as in Figure \ref{1e8}. 
    \begin{figure}
        \includegraphics[width=9cm]{1e8.png} 
        \caption{Comparative Performance with for $10^7$ elements}
        \label{1e8}
    \end{figure}

    This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficeicny gains on larger array sizes. 
    \section{Conclusion}
    % TODO these are what we expect to see...
    As expected, the three $O(n log(n))$ algorithms perform better then Insertion sort. 
    Due to the more parallel nature of Quicksort and Mergesort, these algorithms benefit more from parallization then the more sequential Heapsort. 
    This is because the divide and conquer strategy is inherently more parallel, which should be taken into account when developing new algorithms to be run on parallel and distributed computing platforms. 
    % TODO further findings from implementation
    \bibliography{ref.bib}
    \bibliographystyle{ieeetran}
    \appendix

	All of our code and data can be found in our GitHub repo \href{https://github.com/causeyc/csds438_sorting}{here}. Our implementation of parallel Mergesort is located in mergesort/merge$\_$sort$\_$OpenMP.c. The parallel Quicksort implementation is in quicksort/quicksortOMP.c. Our serial Heapsort implementation is located in heapsort/heapsort.c. In order to run the experiment, clone the repo into a location on Case Western's HPC Markov cluster. From there, run the SLURM batch scripts. The first is \textbf{sorting.slurm} and is located in mergesort/. This script runs parallel Mergesort on a compute node on all combinations of array sizes and CPU core numbers (as explained in the main part of the paper). Once finished, an output SLURM file will be produced that contains the execution times of the sorts. \textit{Mutatis mutandis} for Quicksort (the batch script is called \textbf{qs-job.slurm} and is located in quicksort/). CSV files that we manually created from the data produced from our SLURM jobs are located in mergesort/ and quicksort/ and are called "MergeSort Data.csv" and "project.csv", respectively.

>>>>

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
		T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
	\title{Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP}
	\author{\IEEEauthorblockN{Benjamin Pierce\\\href{mailto:bgp12@case.edu}{bgp12@case.edu}}
		\IEEEauthorblockA{
			\textit{Department of Computer and Data Sciences}\\ }
		\and
		\IEEEauthorblockN{Alberto Safra\\\href{mailto:axs1249@case.edu}{axs1249@case.edu}}
		\IEEEauthorblockA{
			\textit{NCDP, Deans’ Department}}
		\and
		\IEEEauthorblockN{Colin Causey\\\href{mailto:cjc158@case.edu}{cjc158@case.edu}}
		\IEEEauthorblockA{
			\textit{Department of Computer and Data Sciences}}
		\and
		\IEEEauthorblockN{Jason Richards\\\href{mailto:jdr145@case.edu}{jdr145@case.edu}}
		\IEEEauthorblockA{
			\textit{Department of Computer and Data Sciences}}}
	\maketitle

@@

	As mentioned before, all comparison based sorts have an asymptotic lower bound of $\Omega(nlog(n)$.
	Indeed,  Mergesort, and Heapsort achieve a $O(nlog(n))$ upper bound as well, and Quicksort behaves as $O(nlog(n))$ in the average case.
	Insertion sort, as a less advanced algorithm, has an upper bound of $O(n^2)$, as does Quicksort. \cite{cormen_introduction_2009}
	However, Quicksort's upper bound is rarely the case, and tends to have smaller constant factors then either Mergesort or Heapsort. \cite{hoare_algorithm_1961} % TODO wrong cite
	Additionally, each algorithm is constructed differently, and some are more amenable to parallelization than others.

@@

	\subsection{Heapsort}
	The final sort discussed here is Heapsort. Heapsort is yet another $O(n log(n))$ comparison sort.
	It, along with the heap data structure, was invented in 1964. \cite{forsythe_algorithms_1964}
	Heapsort first turns the dataset into a max heap, which is a binary tree where each parent node is greater than its children.
	This process, called \textit{heapification}, is an $O(n)$ algorithm.
	Sorting is performed by repeatedly popping the root node (the maximum value) and re-heapifying.
	This procedure takes advantage of the binary tree structure, and is worst case $O(n log (n))$ overall.  \cite{cormen_introduction_2009}
	Unfortunately, Heapsort is a poor candidate for parallelization, as it does not divide or partition the input array into subarrays and depends on the root node being the absolute maximum.

@@

	In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. \cite{blelloch_programming_1996}
	This is because each subarray can be sorted independently, and this leads to speedup.
	% TODO @Jason add detail on your implementation
	\section{Methodology}
	In this study, the OpenMP \cite{openmp08} API is used for parallelization.
	It is an example of a fork-join methodology, where each parallel thread forks off from a main thread, and then the results are joined back together.
	This is implemented into the C programming language via \textit{\#pragma} preprocessor directives.
	OpenMP is useful for obviously parallel cases, such as Mergesort and Quicksort, as the division of work is quite clear.
	% TODO add more OMP detail here.

@@

	\section{Results}
	In this section, the results of each algorithm will be shown.
	The main variable of interest is how much time each algorithm takes as a function of the length of the input array $n$. In particular, Quicksort is perhaps the most parallelizable sorting algorithm, and can achieve a linear speedup with few modifications. [?] This is because each subarray can be sorted independently, and this leads to speedup. However, in practice, our mergesort implementation is generally the most quick.

@@

	\subsection{Quicksort}
	There was great success with parallel Quicksort.
	As a divide-and-conquer algorithm, it is a naturally parallel algorithm.
	The results of this algorithm can be seen in Figure \ref{qck_per}.
	\begin{figure}[h]
		\includegraphics[width=9.25cm]{qs_mt.png}
		\caption{Performance of Quicksort}
		\label{qck_per}
	\end{figure}

@@

	\begin{algorithm}
		\SetAlgoLined
		\KwResult{A sorted list }
		list $a$\;
		list $result$\;
		int $threads$\;
		let $lists$ be $a$ partitioned into $threads$\\
		\While{every list in lists is not empty }{
			\textbf{In Parallel}\;
			\hspace{0.5cm}$heapify$ each list\;
			$pop$ the largest element of all sub-heaps into $result$\;
			$re-heapify$ the heap that has been popped\;
		}
		\Return $result$
		\caption{Parallel-Heaps}
		\label{alg}
	\end{algorithm}
	Algorithm \ref{alg} presents a possibly parallel algorithm that utilizes the idea of Heapsort.
	Essentially, the algorithm divides the array into many smaller heaps, each managed by a single thread.
	This way, all heapification can be done in parallel, and the "max" item is selected from the number of heaps through a straightforward traversal of the roots of the sub-heaps.
	However, this algorithm has several downsides.
	One is that the end reheapification is still sequential; as by necessity, one heap must be popped, the algorithm must wait on the reheapification of that heap.
	This heap will be smaller by a constant factor then the traditional, single-threaded Heapsort, however.
	Therefore, $Parallel-Heaps$ will be at most a constant speedup for Heapsort.

@@

	For our analysis, we attempted to implement an additional algorithm, Shellsort. We were unsuccessful. We include Shellsort here, however, for completeness and for the Case Western Reserve University HPC developer that may find themselves wondering about Shellsort's parallel implementation. Shellsort is an seen to be a generalization of a more elementary sorting algorithm, Insertionsort, and uses the Insertionsort algorithm in its subroutines. Because of Insertionsort's dependence on access to the entire input array per swap movement, all Insertionsort swaps must be done in serial or risk a data race, which would jeopardize the sort's integrity. For this reason, Insertionsort is nonparallelizable, yet performant in small array size scenarios. Shellsort, however, improves on Insertionsort's design by incorporating a divide and conquer strategy. In particular, Shellsort begins by in-place Insertion-sorting small, distant subsets of the input array before performing Insertionsort on the whole resulting array. This final Insertionsort is done because Insertionsort has fantastic $O(N)$ performance for almost-sorted arrays.  Since all Shellsort's subroutines contribute to advancing the array's state in regards to reaching an optimal setup for the final Insertionsort call and that Shellsort's final serial Insertion sort call on the whole array will inevitably fix any gaps the setup stage missed, Shellsort is a natural candidate to parallelization. A perfunctory parallelization attempt, like the one attempted in the appendix, then can implemented by partitioning an input array into disjoint blocks, having sorting them each sorted in parallel with an OMP recursive call to Shellsort, and then running Shellsort in serial on the result of all those in-place recursive sorts. Note Shellsort's particular potential success in giving each CPU of the OMP tasks significant work to do per unit parallel time. This is what would contribute to such an Implementations potential average span of $O(logn)$. Unfortunately, such an implementation was not found to be possible for us at this time, due to issues with redistributing the number of active cores in a loop using the OpenMP framework.

	\subsection{Comparative Results}
	Results are compared in a few different ways.
	One is to hold the number of threads constant and plot the performance of each algorithm, as in Figure \ref{msqshs}, which shows that Heapsort, which is not parallelized, proves to be much worse than either Quicksort or Mergesort.
	\begin{figure}
		\includegraphics[width=9.25cm]{mshsqs.png}
		\caption{Comparative Performance with 8 threads}
		\label{msqshs}
	\end{figure}
	%TODO Add deets
	Alternatively, it is useful to compare results on a single array size with multiple threads.
	This can be seen for $10^6$ elements in Figure \ref{1e5}
	\begin{figure}
		\includegraphics[width=9cm]{1e5.png}
		\caption{Comparative Performance with for $10^4$ elements}
		\label{1e5}
	\end{figure}
	Interestingly, the performance of Quicksort appears to worsen with the number of threads more than Mergesort; this suggests significant overhead introduced by parallelization.
	However, this effect goes away with an increase in the number of elements, as in Figure \ref{1e8}.
	We can conclude from
	\begin{figure}
		\includegraphics[width=9cm]{1e8.png}
		\caption{Comparative Performance with for $10^7$ elements}
		\label{1e8}
	\end{figure}

	This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficiency gains on larger array sizes. 
	This is a natural consequence of parallel processes, and is a reason why parallelization is not enabled by default. 
	\section{Conclusion}
	% TODO these are what we expect to see...
	Both Mergesort and Quicksort parallelize very efficiently. Based on the trends of Figures 5-8, our parallel Mergesort generally outperforms our parallel Quicksort with respect to our first criterion of evaluation: execution time. This was an interesting result as Quicksort generally has a reputation of outperforming Mergesort in practice. With respect to our second criterion of evaluation, scalability, Mergesort and Quicksort are very similar. They both scale very efficiently as the CPU counts increase, for sufficiently large arrays. According to our criteria of evaluation, therefore, we conclude that with respect to choosing a parallel sorting algorithm to use on Case Western Reserve University’s HPC Markov cluster, our implementation of parallel Mergesort is the preferred algorithm. It should, however, be stated that Mergesort does have a heavier memory requirement as it uses an auxiliary array to accomplish sorting, while Quicksort sorts in-place. With that being said, memory was not a bottleneck as our arrays all fit into main memory. In any case, both parallel Mergesort and parallel Quicksort (as expected) dramatically outperform serial Heapsort, thus demonstrating the great advantages of using parallel computing to efficiently solve the sorting problem.

	% TODO further findings from implementation
	\bibliography{ref.bib}
	\bibliographystyle{ieeetran}
	\appendix

	All of our code and data can be found in our public GitHub repository \href{https://github.com/causeyc/csds438_sorting}{here}. Our implementation of parallel Mergesort is located in mergesort/merge$\_$sort$\_$OpenMP.c. The parallel Quicksort implementation is in quicksort/quicksortOMP.c. Our serial Heapsort implementation is located in heapsort/heapsort.c. In order to run the experiment, clone the repo into a location on Case Western's HPC Markov cluster. From there, run the SLURM batch scripts. The first is \textbf{sorting.slurm} and is located in mergesort/. This script runs parallel Mergesort on a compute node on all combinations of array sizes and CPU core numbers (as explained in the main part of the paper). Once finished, an output SLURM file will be produced that contains the execution times of the sorts. \textit{Mutatis mutandis} for Quicksort (the batch script is called \textbf{qs-job.slurm} and is located in quicksort/). CSV files that we manually created from the data produced from our SLURM jobs are located in mergesort/ and quicksort/ and are called "MergeSort Data.csv'' and "project.csv", respectively. This information is also available in our README.
